CRUSE to Safe Cycling in Ireland \\\large{An Open Source Tool to Support Active Travel}

### 2023-10-09 paper.qmd
Update paper, no footnotes

In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030.[^1]
Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.

[^1]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>

A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
The main frameworks underpinning these efforts are the Climate Action Plan[^2], the National Development Plan[^3], and the National Roads 2040 strategy
[^4].
<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,[^5] "83% of respondents supported enhanced walking and cycling facilitie
s".

[^2]: <https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/>

[^3]: <https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/>

[^4]: https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf

[^5]: https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf

At the regional level, the recently published Greater Dublin Area Transport Strategy[^6] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car[^7]
. Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode"
[^8].

[^6]: <https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/>

[^7]: https://www.nationaltransport.ie/wp-content/uploads/2019/01/National_Household_Travel_Survey_2017_Report\_-\_December_2018.pdf

[^8]: https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/

@@

The methods used to generate the evidence presented in the CRUSE Tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales"[^9] by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[^10]
.

[^9]: See REF Impact Case Study "Cycle network policy, planning and investment transformed by the Propensity to Cycle Tool" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f) submitted by the University of Leeds.

[^10]: See the REF Impact Case Study "Creating Step Changes in Cycling Policy and Infrastructure Planning across the UK" at [results2021.ref.ac.uk](https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B) submitted by the University of Westminster.

@@

The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
In Portugal the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool[^11] and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.

[^11]: See the Health economic assessment tool for walking and cycling at [www.heatwalkingcycling.org](https://www.heatwalkingcycling.org/#homepage).

@@

Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^12]
The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.

[^12]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf

>>>>

In Ireland, the Road Safety Authority (RSA) has set the target of halving the number of road traffic deaths and serious injuries by 2030 [@national2021].
Doing so while simultaneously enabling rapid uptake of active modes will require key active travel routes to be identified and improved.

A proactive approach to the development of cycling policy and infrastructure is taking place in Ireland.
The main frameworks underpinning these efforts are the Climate Action Plan [@climate2022a], the National Development Plan, and the National Roads 2040 strategy [@national2023].
<!-- Strategic alignment is also outlined by the Phoenix Park Transport and Mobility Options Study, which highlights public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016,^[ --> These policy frameworks are backed-up by public support for active travel interventions: while only 3% of trips in Ireland are made by cycling according to the latest Census data from 2016 "83% of respondents supported enhanced walking and cycling facilities" [@sustaina2019].

At the regional level, the recently published Greater Dublin Area Transport Strategy [@greater] reinforces these findings: "nearly a quarter of adults cycle at least once a week in the Dublin Metropolitan Area" with cycling in the Dublin area taking up to 60,000 cars off the road today.
Extrapolating this on a per population basis across Ireland, with around 40% of the population living in Dublin, this suggests that around 150,000 cars could be removed nationwide just by achieving Dublin levels of cycling in all counties (notwithstanding existing cycling trips and differences in trip distances).
7 in 10 trips in Ireland are by car [@national] .
Cycling has the potential to replace a large proportion of these trips: "A high priority must also be given to cyclists, because trips by this mode have the potential to replace trips by private car, most specifically for short to medium distance trips, but increasingly for longer trips as e-bikes extend the range of this mode" [@greater].

@@

The methods used to generate the evidence presented in the CRUSE Tool for Ireland build on the Propensity to Cycle Tool (PCT), which was originally funded by the UK's Department for Transport and developed by a multi-university team.
An important feature of the PCT is that it is open source and publicly available (at [www.pct.bike](https://www.pct.bike/)), allowing its use by all stakeholders in the transport planning process [@lovelace2017].
The PCT approach has had major policy impacts, as outlined in Research Excellence Framework (REF) impact case studies, which demonstrate that the tool "revolutionised strategic cycle planning in England and Wales" by overcoming the barriers to cycling investment imposed by lack of evidence on cycling potential[@lovelace2023].

@@

The PCT approach has been applied in other countries, including Ireland (the topic of this paper), Scotland, and Portugal.
In Portugal the 'biclaR' project, based on methods underlying the PCT, has been developed and deployed for the Lisbon metro region.
The resulting evidence is available in an interactive web application hosted at [biclar.tmlmobilidade.pt](https://biclar.tmlmobilidade.pt) [@felix2023].
biclaR includes estimates of impacts, using the World Health Organisation (WHO) 'HEAT for Cycling' tool and an 'intermodality' scenario that combines cycling with currently available public transit options based on General Transit Feed Specification (GTFS) data.

@@

Although Dublin City has made progress, there are still many parts of the Fastest route network, and even some parts of the Balanced and Quietest route networks, that are not cycle friendly and which have high cycling potential.
According to a recent report, 71% of residents in the Dublin Metropolitan Area support "more cycle tracks along roads, physically separated from traffic and pedestrians".[^1]
The results for the Dublin area could help prioritise investment in cycling infrastructure in the city.

[^1]: https://www.nationaltransport.ie/wp-content/uploads/2022/05/220504-WACI22_DublinMetropolitanArea_v35_DIGITAL_v2.pdf

### 2023-10-09 paper.tex
Update paper, no footnotes

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, interior hidden, breakable, boxrule=0pt, enhanced, sharp corners, frame hidden]}{\end{tcolorbox}}\fi

@@

Uptake of cycling should be seen in this broader context of transport
decarbonisation \citep{brand2020} and sustainable mobility
\citep{burns2013}. Although cycling uptake appears on the surface to
only relate to the `shift' part of the ASI framework, closer
consideration of the knock-on impacts of cycling uptake shows that it
can also help avoid unnecessary trips \citep{nello-deakin2020}.
Furthermore, highly efficient e-bikes --- which are seeing rapid uptake
in Ireland and other countries --- outperform electric cars, which are
too heavy and expensive. Over-reliance on electric cars could slow the
transition away from car dependency and inadvertently enable ``high
travel lock-in'' \citep{anable2019}. At the European level, the European
Union has a target of reducing GHG emissions by 55\% by 2030, compared
to 1990 levels, and to achieve `net-zero' by 2050 \citep{directiv2008}.

Regarding road traffic casualties, another deadly consequence of
inefficient transport systems, the Road Infrastructure Safety Management
(RISM) directive (2008/96/EC) requires member states to implement a road
safety management system (RSMS) for all public roads. Specifically,
``Member States shall ensure that the ranking of high accident
concentration sections and the network safety ranking are carried
out''.\footnote{https://eur-lex.europa.eu/legal-content/EN/TXT/?u
  ri=celex\%3A32008L0096} Given that `safety' in this context is
usefully quantified as the number of people killed and seriously injured
(KSI) per distance travelled, the directive requires estimation of
distance travelled by mode, down to the road link level. For active
modes, about which there is a paucity of data compared with motorized
modes, this is a major challenge. Better data to inform road safety
policies and interventions is a motivation for the CRUSE Tool outlined
in this paper, and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by
2030.\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}}
Doing so while simultaneously enabling rapid uptake of active modes will
require key active travel routes to be identified and improved.

A proactive approach to the development of cycling policy and
infrastructure is taking place in Ireland. The main frameworks
underpinning these efforts are the Climate Action Plan\footnote{\url{https://www.gov.ie/en/publication/7bd8c-climate-action-plan-2023/}},
the National Development Plan\footnote{\url{https://www.gov.ie/en/publication/774e2-national-development-plan-2021-2030/}},
and the National Roads 2040 strategy \footnote{https://www.tii.ie/tii-library/strategic-planning/national-roads-2040/TII-NR2040-Final-Report-EN-April-2023.pdf}.
These policy frameworks are backed-up by public support for active
travel interventions: while only 3\% of trips in Ireland are made by
cycling according to the latest Census data from 2016,\footnote{https://assets.gov.ie/42950/f789b2fcfb144441822f58236986eeea.pdf}
``83\% of respondents supported enhanced walking and cycling facilitie
s''.

At the regional level, the recently published Greater Dublin Area
Transport Strategy\footnote{\url{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}}
reinforces these findings: ``nearly a quarter of adults cycle at least
once a week in the Dublin Metropolitan Area'' with cycling in the Dublin
area taking up to 60,000 cars off the road today. Extrapolating this on
a per population basis across Ireland, with around 40\% of the
population living in Dublin, this suggests that around 150,000 cars
could be removed nationwide just by achieving Dublin levels of cycling
in all counties (notwithstanding existing cycling trips and differences
in trip distances). 7 in 10 trips in Ireland are by car\footnote{https://www.nationaltransport.ie/wp-content/uploads/2019/01/National\_Household\_Travel\_Survey\_2017\_Report\_-\_December\_2018.pdf}
. Cycling has the potential to replace a large proportion of these
trips: ``A high priority must also be given to cyclists, because trips
by this mode have the potential to replace trips by private car, most
specifically for short to medium distance trips, but increasingly for
longer trips as e-bikes extend the range of this mode'' \footnote{https://www.nationaltransport.ie/planning-and-investment/strategic-planning/greater-dublin-area-transport-strategy/}.

@@

The methods used to generate the evidence presented in the CRUSE Tool
for Ireland build on the Propensity to Cycle Tool (PCT), which was
originally funded by the UK's Department for Transport and developed by
a multi-university team. An important feature of the PCT is that it is
open source and publicly available (at
\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
stakeholders in the transport planning process \citep{lovelace2017}. The
PCT approach has had major policy impacts, as outlined in Research
Excellence Framework (REF) impact case studies, which demonstrate that
the tool ``revolutionised strategic cycle planning in England and
Wales''\footnote{See REF Impact Case Study ``Cycle network policy,
  planning and investment transformed by the Propensity to Cycle Tool''
  at
  \href{https://results2021.ref.ac.uk/impact/847d1191-7f25-46ba-a399-b481125edc8f}{results2021.ref.ac.uk}
  submitted by the University of Leeds.} by overcoming the barriers to
cycling investment imposed by lack of evidence on cycling
potential\footnote{See the REF Impact Case Study ``Creating Step Changes
  in Cycling Policy and Infrastructure Planning across the UK'' at
  \href{https://results2021.ref.ac.uk/impact/4BBF3436-FD10-4C75-9791-F5E98AB4411B}{results2021.ref.ac.uk}
  submitted by the University of Westminster.} .

@@

The PCT approach has been applied in other countries, including Ireland
(the topic of this paper), Scotland, and Portugal. In Portugal the
`biclaR' project, based on methods underlying the PCT, has been
developed and deployed for the Lisbon metro region. The resulting
evidence is available in an interactive web application hosted at
\href{https://biclar.tmlmobilidade.pt}{biclar.tmlmobilidade.pt}
\citep{felix2023}. biclaR includes estimates of impacts, using the World
Health Organisation (WHO) `HEAT for Cycling' tool\footnote{See the
  Health economic assessment tool for walking and cycling at
  \href{https://www.heatwalkingcycling.org/\#homepage}{www.heatwalkingcycling.org}.}
and an `intermodality' scenario that combines cycling with currently
available public transit options based on General Transit Feed
Specification (GTFS) data.

>>>>

\begin{keyword}
    Cycling \sep Open-source \sep Road Safety \sep 
    Active Travel
\end{keyword}
\end{frontmatter}
    \ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[sharp corners, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, enhanced, interior hidden, breakable, frame hidden]}{\end{tcolorbox}}\fi

@@

Uptake of cycling should be seen in this broader context of transport
decarbonisation \citep{brand2020} and sustainable mobility
\citep{burns2013}. Although cycling uptake appears on the surface to
only relate to the `shift' part of the ASI framework, closer
consideration of the knock-on impacts of cycling uptake shows that it
can also help avoid unnecessary trips \citep{nello-deakin2020}.
Furthermore, highly efficient e-bikes --- which are seeing rapid uptake
in Ireland and other countries --- outperform electric cars, which are
too heavy and expensive. Over-reliance on electric cars could slow the
transition away from car dependency and inadvertently enable ``high
travel lock-in'' \citep{anable2019}. At the European level, the European
Union has a target of reducing GHG emissions by 55\% by 2030, compared
to 1990 levels, and to achieve `net-zero' by 2050 \citep{rosenow2022}.

Regarding road traffic casualties, another deadly consequence of
inefficient transport systems, the Road Infrastructure Safety Management
(RISM) directive (2008/96/EC) requires member states to implement a road
safety management system (RSMS) for all public roads. Specifically,
``Member States shall ensure that the ranking of high accident
concentration sections and the network safety ranking are carried out''
\citep{directiv2008}. Given that `safety' in this context is usefully
quantified as the number of people killed and seriously injured (KSI)
per distance travelled, the directive requires estimation of distance
travelled by mode, down to the road link level. For active modes, about
which there is a paucity of data compared with motorized modes, this is
a major challenge. Better data to inform road safety policies and
interventions is a motivation for the CRUSE Tool outlined in this paper,
and modeling active travel more broadly.

In Ireland, the Road Safety Authority (RSA) has set the target of
halving the number of road traffic deaths and serious injuries by 2030
\citep{national2021}. Doing so while simultaneously enabling rapid
uptake of active modes will require key active travel routes to be
identified and improved.

A proactive approach to the development of cycling policy and
infrastructure is taking place in Ireland. The main frameworks
underpinning these efforts are the Climate Action Plan
\citep{climate2022a}, the National Development Plan, and the National
Roads 2040 strategy \citep{national2023}. These policy frameworks are
backed-up by public support for active travel interventions: while only
3\% of trips in Ireland are made by cycling according to the latest
Census data from 2016 ``83\% of respondents supported enhanced walking
and cycling facilities'' \citep{sustaina2019}.

At the regional level, the recently published Greater Dublin Area
Transport Strategy \citep{greater} reinforces these findings: ``nearly a
quarter of adults cycle at least once a week in the Dublin Metropolitan
Area'' with cycling in the Dublin area taking up to 60,000 cars off the
road today. Extrapolating this on a per population basis across Ireland,
with around 40\% of the population living in Dublin, this suggests that
around 150,000 cars could be removed nationwide just by achieving Dublin
levels of cycling in all counties (notwithstanding existing cycling
trips and differences in trip distances). 7 in 10 trips in Ireland are
by car \citep{national} . Cycling has the potential to replace a large
proportion of these trips: ``A high priority must also be given to
cyclists, because trips by this mode have the potential to replace trips
by private car, most specifically for short to medium distance trips,
but increasingly for longer trips as e-bikes extend the range of this
mode'' \citep{greater}.

@@

The methods used to generate the evidence presented in the CRUSE Tool
for Ireland build on the Propensity to Cycle Tool (PCT), which was
originally funded by the UK's Department for Transport and developed by
a multi-university team. An important feature of the PCT is that it is
open source and publicly available (at
\href{https://www.pct.bike/}{www.pct.bike}), allowing its use by all
stakeholders in the transport planning process \citep{lovelace2017}. The
PCT approach has had major policy impacts, as outlined in Research
Excellence Framework (REF) impact case studies, which demonstrate that
the tool ``revolutionised strategic cycle planning in England and
Wales'' by overcoming the barriers to cycling investment imposed by lack
of evidence on cycling potential\citep{lovelace2023}.

@@

The PCT approach has been applied in other countries, including Ireland
(the topic of this paper), Scotland, and Portugal. In Portugal the
`biclaR' project, based on methods underlying the PCT, has been
developed and deployed for the Lisbon metro region. The resulting
evidence is available in an interactive web application hosted at
\href{https://biclar.tmlmobilidade.pt}{biclar.tmlmobilidade.pt}
\citep{felix2023}. biclaR includes estimates of impacts, using the World
Health Organisation (WHO) `HEAT for Cycling' tool and an `intermodality'
scenario that combines cycling with currently available public transit
options based on General Transit Feed Specification (GTFS) data.
